### 2024-08-27 teaching/math240f24/notes.tex
Notes draft

\input fontmac
\input mathmac
\widemargins
\bookheader{\tenpoint\sc marcel goh}{\tenpoint\sc math {\oldstyle 240} fall {\oldstyle 2024}}

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}

@@

\floattext 4.5 \ninebf A note on these notes.
\ninepoint After each class, this document will be updated with the new material that was
just covered. The datestamps in the left margin indicate when the notes from each day start.
Subsections labelled with a $*$ are optional. This document is based on notes by
Jeremy Macdonald, but any errors are likely my own. Please email me if you find any.

@@

Next, we define the {\it difference} $B\setminus A$ of two sets. This is the set of all elements
in $B$ that are {\it not} in $A$. So, using the complement notation we just learned about,
we can express $B\setminus A = B \cap \bar A$. It is not necessary that $A$ be a subset of $B$. In
the small example above, we have $A\setminus B = \{2,4\}$ and $B\setminus A = \{3,5\}$.

Lastly, we define the {\it symmetric difference} $A \symdiff B$ of two sets $A$ and $B$ to be the set of all
elements that are either in $A$ or in $B$ {\it but not both}. Invoking the above example one last time, we
have $A\symdiff B = \{2,3,4,5\}$. To practise using all the different operations we just learned,
convince yourself that the following are all valid ways to express the symmetric difference:
$$A\symdiff B = (A\cup B)\setminus (A\cap B) = (A\setminus B) \cup (B\setminus A) =
\bar{\bar{A\cup B} \cup (A\cap B)}$$
\datestamp{2}{} More set identities abound. We state the following proposition without proof; you should try going
though this list and convincing yourself that each identity holds, for all sets $A$, $B$, and $C$.
(This is a great way of practising proofs by double inclusion.)

\proclaim Proposition 2. Let $A$, $B$, and $C$ be subsets of a universe $U$. Then
\medskip
\item{i)} $A\cap U = A$ and $A\cup \emptyset = A$;
\smallskip
\item{ii)} $A\cup U = U$ and $A\cap \emptyset = \emptyset$;
\smallskip
\item{iii)} $A\cup A = A$ and $A\cap A = A$;
\smallskip
\item{iv)} $\bar{\bar A} = A$;
\smallskip
\item{v)} $A\cup B = B\cup A$ and $A\cap B = B\cap A$;
\smallskip
\item{vi)} $A\cup (B\cup C) = (A\cup B) \cup C$ and $A\cap (B\cap C) = (A\cap B)\cap C)$;
\smallskip
\item{vii)} $A\cup (B\cap C) = (A\cup B) \cap (A\cup C)$ and $A\cap (B\cup C) = (A\cap B)\cup (A\cap C)$;
\smallskip
\item{viii)} $A\cup (A\cap B) = A$ and $A\cap (A\cup B) = A$; and
\smallskip
\item{ix)} $A\cup \bar A = U$ and $A\cap \bar A = \emptyset$.\slug
\medskip

Some of these laws have names: (v) is called the commutative law, (vi) is called the associative
law, (vii) is the distributive law, (viii) is called the absorption law, and (ix) is the complement law.

@@

A {\it proposition} is a statement that is true or false. For example ``$8$ is even'' is a statement
we know to be true, and ``$8$ is prime'' is a statement we know to be false. The statement
``$n$ is prime'' is not a proposition because its truth or falsity depends on what $n$ is. (But if
$n$ is previously fixed to stand for some value, then this would be a proposition.) The statement
``$2^{2^240} - 1$ is prime'' is a proposition, because it is either true or false (even though you or I
might not know which one it is).

A {\it propositional variable} or a {\it boolean variable} is a variable which can take either the
value $0$ or $1$, where $0$ means ``false'' and $1$ means ``true.'' Usually we use letters $p$, $q$,
and $r$ to denote propositional variables.

>>>>

\input fontmac
\input mathmac

\input epsf
\widemargins
\bookheader{\tenpoint\sc marcel goh}{\tenpoint\sc math {\oldstyle 240} fall {\oldstyle 2024}}

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}
\def\To{\Rightarrow}
\def\Gets{\Leftarrow}

@@

\floattext 4.5 \ninebf A note on these notes.
\ninepoint After each class, this document will be updated with the new material that was
just covered. The datestamps in the left margin indicate when the notes from each day start.
Subsections labelled with a $*$ are optional. This document is heavily based on notes by
Jeremy Macdonald, but any errors are likely my own. Please email me if you find any.

@@

Next, we define the {\it difference} $B\setminus A$ (sometimes $B-A$) of two sets.
This is the set of all elements
in $B$ that are {\it not} in $A$. So, using the complement notation we just learned about,
we can express $B\setminus A = B \cap \bar A$. It is not necessary that $A$ be a subset of $B$. In
the small example above, we have $A\setminus B = \{2,4\}$ and $B\setminus A = \{3,5\}$.

Lastly, we define the {\it symmetric difference} $A \symdiff B$ of two sets $A$ and $B$ to be the set of all
elements that are either in $A$ or in $B$ {\it but not both}.
Invoking the above example one last time, we
have $A\symdiff B = \{2,3,4,5\}$.
We can express it as using unions, intersections,
and complements as
\edef\eqsymdiffdef{\the\eqcount}
$$A \symdiff B = (A\cup B) \cap \bar{A \cap B}.\adveq$$
To practise using all the different operations we just learned,
convince yourself that the following are three more valid ways to express the symmetric difference:
$$A\symdiff B = (A\cup B)\setminus (A\cap B) = (A\setminus B) \cup (B\setminus A) =
\bar{\bar{A\cup B} \cup (A\cap B)}$$
\datestamp{2}{} More set identities abound. We state the following proposition without proof; you should try going
though this list and convincing yourself that each identity holds, for all sets $A$, $B$, and $C$.
(This is a great way of practising proofs by double inclusion.)

\edef\propsetlaws{\the\thmcount}
\proclaim Proposition \advthm. Let $A$, $B$, and $C$ be subsets of a universe $U$. Then
\medskip
\item{i)} $A\cap U = A$ and $A\cup \emptyset = A$;
\smallskip
\item{ii)} $A\cup U = U$ and $A\cap \emptyset = \emptyset$;
\smallskip
\item{iii)} $A\cup A = A$ and $A\cap A = A$;
\smallskip
\item{iv)} $\bar{\bar A} = A$;
\smallskip
\item{v)} $A\cup B = B\cup A$ and $A\cap B = B\cap A$;
\smallskip
\item{vi)} $A\cup (B\cup C) = (A\cup B) \cup C$ and $A\cap (B\cap C) = (A\cap B)\cap C)$;
\smallskip
\item{vii)} $A\cup (B\cap C) = (A\cup B) \cap (A\cup C)$ and $A\cap (B\cup C) = (A\cap B)\cup (A\cap C)$;
\smallskip
\item{viii)} $A\cup (A\cap B) = A$ and $A\cap (A\cup B) = A$; and
\smallskip
\item{ix)} $A\cup \bar A = U$ and $A\cap \bar A = \emptyset$.\slug
\medskip

These laws have names, some of which we will use more often than others:
(i) is called the identity law, (ii) the domination law, (iii) the idempotent law,
(iv) the law of double negation, (v) the commutative law, (vi) the associative law,
(vii) the distributive law, (viii) the absorption law, and (ix) the complement law.

@@

A {\it proposition} is a statement that is true or false. For example ``$8$ is even'' is a statement
we know to be true, and ``$8$ is prime'' is a statement we know to be false. The statement
``$n$ is prime'' is not a proposition because its truth or falsity depends on what $n$ is.
The statement
``$2^{2^{240}} - 1$ is prime'' is a proposition, because it is either true or false (even though you or I
might not know which one it is).

A {\it propositional variable} or a {\it boolean variable} is a variable which can take either the
value $0$ or $1$, where $0$ means ``false'' and $1$ means ``true.'' Usually we use letters $p$, $q$,
and $r$ to denote propositional variables. The simplest logical operator is negation, defined
by the table
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil \cr
p & \neg p \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 1 \cr
1 & 0 \cr
}}.
$$
This is also called the {\mc NOT} operator, since if $p$ is true, then $\neg p$
is false, and vice versa.
Next is {\it conjunction}, which has the table
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\wedge q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 0 \cr
0 & 1 & 0 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}.
$$
This is also called the {\mc AND} operator, because $p\wedge q$ is true if and only if $p$ and
$q$ are both true. The {\mc OR} operator, also called {\it disjunction}, has the table
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\vee q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 0 \cr
0 & 1 & 1 \cr
1 & 0 & 1 \cr
1 & 1 & 1 \cr
}}.
$$
We see that $p\vee q$ is true if $p$ or $q$ is true (or both). The symbol $\vee$ is meant to recall
the Latin word {\it vel}, meaning ``or.'' (One of the most important early treatises on mathematical
logic and set theory was {\sl Arithmetices principia, nova methodo exposita}, published in Latin in 1889
by Giuseppe Peano. It established the now-standard axiomatisation of the natural numbers.)

On the other hand, in English, we often use the word ``or''
to mean an {\it exclusive} or; that is, either $p$ and $q$ are true but not both. In mathematics, on the
other hand, ``or'' is usually {\it inclusive}, so both $p$ and $q$ are allowed to hold at the same time.
It is possible to express the exclusive disjunction (often called {\mc XOR})
by a table, however. We will use the symbol $\oplus$ for this operator, and its table looks like this:
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\oplus q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 0 \cr
0 & 1 & 1 \cr
1 & 0 & 1 \cr
1 & 1 & 0 \cr
}}
$$
So $p\oplus q$ is true if $p$ is true or $q$ is true, but not both.

A {\it formula} is an expression containing propositional variables, $0$, $1$, logical operators, and
parentheses. The formula must syntactically make sense; for instance, $0 (\vee \wedge q \neg $ is
not a formula. Just as in ordinary mathematical notation,
parentheses are used to clarify which operators should be evaluated first. We will assume that negation
applies first, but an expression such as $p\vee q \wedge r$ is ambiguous. (In many programming languages,
conjunction has higher priority than disjunction, but in this class, just add parentheses to clarify.)

Above we have illustrated the basic logical operators
by writing out their {\it truth tables}. These are tables that give the value of
a formula for all possible values of its variables. We can write truth tables for more complex formulas as
well:
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\wedge q & \neg(p \wedge q) & \neg(p\wedge q) \oplus q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 0 & 1 & 1 \cr
0 & 1 & 0 & 1 & 0 \cr
1 & 0 & 0 & 1 & 1 \cr
1 & 1 & 1 & 0 & 1 \cr
}}
$$
Strictly speaking, the third and fourth columns are not necessary, but these intermediary columns help us
verify the accuracy of the following ones.

Two formulas $f_1$ and $f_2$ are said to be {\it logically equivalent} if they have the same truth table;
that is, they produce the same output if given the same input. In this case we write $f_1 \equiv f_2$.
For example, let $f_1 = \neg(p\wedge q) \oplus q$, the formula whose truth table is illustrated above.
Now let $f_2 = p \vee \neg q$. Its truth table is
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & \neg q & p\vee \neg q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 & 1 \cr
0 & 1 & 0 & 0 \cr
1 & 0 & 1 & 1 \cr
1 & 1 & 0 & 1 \cr
}},
$$
so we conclude that $f_1 \equiv f_2$. As a larger example, suppose we want to find all
values of $p$, $q$, and $r$ such that
$$ f = (p\vee q)\wedge (\neg q \vee \neg r)$$
evaluates to $1$. The truth table
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil &
\hfil$#$\hfil & \hfil$#$\hfil  \cr
p & q & r & p\vee q & \neg q \vee \neg r & f \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 0 & 0 & 1 & 0 \cr
0 & 0 & 1 & 0 & 1 & 0 \cr
0 & 1 & 0 & 1 & 1 & 1 \cr
0 & 1 & 1 & 1 & 0 & 0 \cr
1 & 0 & 0 & 1 & 1 & 1 \cr
1 & 0 & 1 & 1 & 1 & 1 \cr
1 & 1 & 0 & 1 & 1 & 1 \cr
1 & 1 & 1 & 1 & 0 & 0 \cr
}}
$$
shows that $f$ evaluates to $1$ precisely when
$$(p,q,r) \in \bigl\{(0,1,0), (1,0,0), (1,0,1), (1,1,0)\bigr\}.$$
To write out the truth table for a formula with $n$ variables, we need $2^n$ rows, so this
method is unsuitable for formulas with more than three or four variables.

\medskip\boldlabel Simplifying logical formulas. Just as we have rules for simplifying set expressions,
there are ways to turn complicated logical formulas into simpler ones that are logically equivalent.
What might surprise you is that the rules turn out to be exactly the same! To see the basis for this
correspondence, consider the definition of a union of sets $A$ and $B$. In set-builder notation,
this is
$$A \cup B = \{ x\in U : x\in A\ \hbox{or}\ x\in B\}.$$
The ``or'' in the definition suggests that $\cup$ is intimately related to the $\vee$ operation in
propositional logic. Repeating this process, we have the ``dictionary''
$$
\vbox{
\tabskip=3em plus.005em minus .005em
\halign{
\hfil{\ninepoint #}\hfil & \hfil{\ninepoint #}\hfil \cr
Set theory & Propositional logic\cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
sets $A$, $B$ & variables $p$, $q$ \cr
unions $A\cup B$ & disjunctions $p\vee q$ \cr
intersections $A\cap B$ & conjunctions $p\wedge q$ \cr
complements $\bar A$ & negations $\neg p$ \cr
symmetric differences $A\symdiff B$ & exclusive disjunctions $p\oplus q$ \cr
the empty set $\emptyset$ & $0$ \cr
the universe $U$ & $1$ \cr
}}
$$
Exploiting this connection, we have the following analogue of Proposition~{\propsetlaws}.

\proclaim Proposition 2. Let $p$, $q$, and $r$ be propositional variables. Then
\medskip
\item{i)} $p\wedge 1 \equiv p$ and $p\vee 0 \equiv p$;
\smallskip
\item{ii)} $p \vee 1 \equiv 1$ and $p \wedge 0 \equiv 0$;
\smallskip
\item{iii)} $p \vee p \equiv p$ and $p\wedge p \equiv p$;
\smallskip
\item{iv)} $\neg\neg p \equiv p$;
\smallskip
\item{v)} $p \vee q \equiv q \vee p$ and $p\wedge q \equiv q\wedge p $;
\smallskip
\item{vi)} $p\vee (q\vee r) \equiv (p\vee q) \vee r$ and $p\wedge (q\wedge r) \equiv (p\wedge q)\wedge r)$;
\smallskip
\item{vii)} $p\vee (q\vee r) \equiv (p\vee q) \wedge (p\vee r)$ and
$p\wedge (q\vee r) \equiv (p\wedge q)\vee (p\wedge r)$;
\smallskip
\item{viii)} $p\vee (p\wedge q) \equiv p$ and $p\wedge (p\vee q) \equiv p$; and
\smallskip
\item{ix)} $p\cup \bar p \equiv 1$ and $p\wedge \bar p \equiv 0$.\slug
\medskip

Since they are essentially the same as their set equivalents, the names of these laws are the same as in the
realm of sets.
We also have the propositional equivalent of De Morgan's law, which states that
\medskip\begingroup\sl
\item{x)} $\neg(p\vee q) \equiv \neg p \wedge \neg q$ and $\neg(p\wedge q) \equiv \neg p \vee \neg q$.\endgroup%sl
\medskip

Using these rules, we can now show that $\neg(p\wedge q) \oplus q \equiv p\vee\neg q$, which we
already saw earlier from their truth tables. First we observe that
\edef\eqxordef{\the\eqcount}
$$p\oplus q \equiv (p\vee q)\wedge \neg(p\wedge q),\adveq$$
which is analogous to the identity~\refeq{\eqsymdiffdef} for symmetric differences. So
$$\eqalign{
\neg(p\wedge q)\oplus q&\equiv\bigl(\neg(p\wedge q)\vee q\bigr)\wedge\neg\bigl(\neg(p\wedge q)\wedge q\bigr)\cr
&\equiv (\neg p \vee\neg q \vee q) \wedge \bigl((p\wedge q) \vee \neg q\bigr) \cr
&\equiv (\neg p \vee 1) \wedge \bigl((p\vee \neg q) \wedge (q\vee \neg q)\bigr) \cr
&\equiv 1 \wedge \bigl((p\vee \neg q) \wedge 1\bigr) \cr
&\equiv p\vee \neg q, \cr
}$$
where in the first line we use~\refeq{\eqxordef}, in the second line we use De Morgan's law twice, in
the third line we use the complement and distributive laws, the fourth line we use the domination and
complement laws, and in the last line we use the identity law (twice). (It is not super important
to remember the names of all these laws, as long as you remember their statements, but you may actually find
it is easier to remember the names along with the statements, just as it might be easier to remember faces of
people you've met if you also know their names.)

\medskip\boldlabel\datestamp{4}{} Conditional and biconditional. We now examine the {\it conditional}
logical relation {\mc IF} $p$ {\mc THEN} $q$. It is denoted by $p\To q$ and its truth table is given by
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\To q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 \cr
0 & 1 & 1 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}.
$$
Within the conditional statement, $p$ is called the {\it antecedent}; this is the assumption. The
statement that is asserted, conditional that the antecendent holds, is called the {\it consequent} $q$.
You can quickly check that the relation $p\To q$ is equivalent to $\neg p \vee q$. This fact is useful
when performing mechanical simplifications.
In English, the statement ``if $p$ then $q$'' asserts a causal relation between $p$ and $q$.
Take a moment and reconcile this idea with the truth table above.
It is normal to get a little tripped up if it's your first time seeing this table.
In the first row, $q$ doesn't even happen, so it might feel weird to say that $p$
has ``caused'' $q$ to happen in this case, and
in the second row, $p$ is not true
and $q$ is true, so it might seem odd that we have set $p\To q$ to true, because there seems to be no relation
between $p$ and $q$.
But it should make sense if you think
of $p$ as a precondition to a promise $q$, and then considering whether the promise is broken. As an example,
suppose your friend says: ``If it snows tomorrow I'll work in Trottier with you.'' If it doesn't snow
and she doesn't pull up, she hasn't technically broken her promise. If it doesn't
snow and she shows up, then she still hasn't broken her promise. The only way she can break her promise is if
it snows and she doesn't come to Trottier; this situation corresponds to the only $0$-row in the truth table.

The conditional is a very important logical operator to understand, because most theorem statements assume
some hypothesis and claim some conclusion. You will be asked to prove statements of this form, so it is
important to understand the logical nature of the statements to begin with.

The last relation is called the {\it biconditional}, and it asserts that variables $p$ and $q$ are logically
equivalent. That is, $p$ happens if and only if $q$ happens. It's truth table
$$
\vbox{
\tabskip=1em plus.2em minus .2em
\halign{
\hfil$#$\hfil & \hfil$#$\hfil & \hfil$#$\hfil \cr
p & q & p\Leftrightarrow q \cr
\noalign{\medskip}
\noalign{\hrule}
\noalign{\medskip}
0 & 0 & 1 \cr
0 & 1 & 0 \cr
1 & 0 & 0 \cr
1 & 1 & 1 \cr
}}
$$
is pretty self-explanatory; in it, $p\Leftrightarrow q$ is true whenever $p$ and $q$ have the same truth value.
The name ``biconditional'' is suggested by (part of) the following proposition.

\proclaim Proposition {\advthm}. Let $p$ and $q$ be propositional variables. Then
$$\eqalign{
p \Leftrightarrow q &\equiv (p\To q) \wedge (q\To p) \cr
&\equiv \bigl( (\neg p)\vee q\bigr) \wedge \bigl( (\neg q)\vee p\bigr) \cr
&\equiv (p\wedge q) \vee \bigl( (\neg p) \wedge (\neg q)\bigr).\cr
}$$

\proof We leave the first equivalence to the reader (writing out the truth table is one way of proving it).
The second equivalence follows from our earlier observation that $p\To q \equiv \neg p \vee q$, and
the third equivalence follows from the distributive, complement, and identity laws.\slug

A formula $f$ is called a
\medskip
\item{i)} {\it tautology} if $f\equiv 1$, i.e., $f$ always evaluates to true;
\smallskip
\item{ii)} {\it contradiction} if $f\equiv 0$, i.e., $f$ always evaluates to false;
\smallskip
\item{iii)} {\it contingency} if $f$ can evaluate to both $1$ and $0$, depending on the
values of its variables;
\smallskip
\item{iv)} {\it satisfiable} if $f$ evaluates to $1$ for at least one input; and
\smallskip
\item{v)} {\it falsifiable} if $f$ evaluates to $0$ for at least one input.
\medskip
An example of a tautology is $p\vee\neg p$ and an example of a contradiction is $p\wedge \neg p$.
This follows from the complement laws. To say that something is satisfiable is precisely to
say that it is not a contradiction, and to say that something is falsifiable is equivalent to saying
that it is not a tautology. Contingencies are those formulas that are both satisfiable
and falsifiable (in other words, formulas that are neither tautologies nor contradictions).

Suppose we are asked which of the above definitions the formula
$$f \equiv \bigl( p\wedge (p\To q)\bigr) \To q$$
satisfies. This formula only has two variables, so it is easy enough to use a truth table
for this purpose, but we will take the opportunity to practise simplifying the expression
symbolically. First of all, let's change all conditionals of the form $r \To s$,
to disjunctions of the form $\neg r \vee s$. This gives us
$$f\equiv \neg\bigl( p\wedge (\neg p \vee q)\bigr) \vee q.$$
Now we use the distributive law to distribute the conjunction over the innermost disjunction,
obtaining
$$f\equiv \neg\bigl((p\wedge \neg p) \vee (p\wedge q)\bigr) \vee q;$$
by the complement and identity laws in that order, this simplifies to
$$f\equiv \neg(p\wedge q) \vee q.$$
Now De Morgan's law and associativity give
$$f\equiv (\neg p \vee \neg q) \vee q \equiv \neg p \vee (\neg q \vee q),$$
and thus
$$f\equiv \neg p \vee 1 \equiv 1,$$
by the complement and domination laws in that order. We conclude that $f$ is a tautology,
which also means that it is satisfiable.

The fact that $\bigl( p\wedge (p\To q)\bigr) \To q$ is a tautology symbolically justifies
the argument that of $p$ is true and $p\To q$ is true, then we should be able to conclude $q$.
This form of argument is called {\it modus ponens}, and it dates back to ancient times. You probably
use {\it modus ponens} all the time in everyday life without knowing it, and we will
certainly use it in this class a lot.

\medskip\boldlabel Encoding problems in propositional logic.
Many algorithmic and logical problems can be encoded in propositional logic (and then later
solved by a computer program). For example, suppose we want to play $4\times 4$ Sudoku. In this
game, we have a $4\times 4$ grid and we want to fill it with the numbers $1$ through $4$ such that
\medskip
\item{i)} every row contains $1$ through $4$;
\smallskip
\item{ii)} every column contains $1$ through $4$; and
\smallskip
\item{iii)} the four subsquares each contain $1$ through $4$.
\medskip
In a given instance of the game, some cells are already filled in. The puzzle is: {\sl Is there a
solution and if so, what is it?}
\midinsert
$$\epsfbox{sudoku.ps}$$
\vskip5pt
\caption{An example $4\times 4$ Sudoku game.}
\endinsert
\goodbreak
To represent a Sudoku game in propositional logic, we can define boolean variables
$p_{i,j,k}$, where $i$, $j$, and $k$ range over $\{1,2,3,4\}$. (So there are $4^3$ variables
in total.) We shall set
$$ p_{i,j,k} = \cases{1, & if the number $k$ is in row $i$ and column $j$;\cr 0, & otherwise.}$$
We'll  number the rows increasing from the top and the columns increasing
from left to right.
For example, in Fig.~1 there is a $2$ in row $1$ and column $3$, so $p_{1,3,2} = 1$.

Now we set to work encoding the conditions of a Sudoku grid in propositional logic:
\medskip
\item{i)} To stipulate that every row contain $1$ through $4$, we first define auxiliary variables
$$r_{i,k} = p_{i,1,k} \vee p_{i,2,k} \vee p_{i,3,k} \vee p_{i,3,k},$$
for $i,k\in \{1,2,3,4\}$.
With these helper variables, we now see that
$$ r_{1,1} \wedge r_{1,2} \wedge r_{1,3} \wedge r_{1,4} $$
encodes the requirement that row $1$ contains one of each number.
We do the same for rows $2$, $3$, and $4$ as well, and then combine with {\mc AND}.
\smallskip
\item{ii)} We do a similar thing as in part (i) for each of the four columns.
\smallskip
\item{iii)} Ditto for subsquares.
\smallskip
\item{iv)} We need to set the initial values of the grid. For the grid in Fig.~1, we have
the formula
$$p_{1,3,2} \wedge p_{3,1,1} \wedge p_{3,2,3}.$$
\smallskip
\item{v)} Lastly, we need to make sure that there is not more than one number per cell. To do
this, for each cell $(i,j)\in \{1,2,3,4\}^2$ we write
$$ p_{i,j,1} \To (\neg p_{i,j,2} \wedge \neg p_{i,j,3} \wedge \neg p_{i,j,4}),$$
and so on (four conditionals in total). Of course we'll need to {\mc AND} all these together.
\medskip
Now we combine the formulas from each of these five steps into one long formula $f$ such that
$f$ is satisfiable if and only if the grid has a solution, and the values for $p_{i,j,k}$ give
a solution.

Defining all of these variables was a rather arduous and cumbersome process, and not entirely worth
it for a $4\times 4$ game of Sudoku (which can just be solved by eyeballing the grid).
But one could imagine writing a general computer program to encode larger and larger grids.
In fact, there are lots of problems that can be {\it reduced} to the problem of
determining if a boolean formula is satisfiable. This means that if we have a program
capable of taking a formula $f$ as input and spitting out whether or not it is satisfiable (in a
reasonable amount of time), then there are lots of real-world problems that this program
could be applied to.

This problem is called the {\it boolean satisfiability problem}, often abbreviated {\mc SAT}. One
way of solving it for any given $f$ is to just compute its truth table. We already know the
downside of this approach: if $f$ has $n$ variables, then its truth table will have $2^n$ rows.
Given a few minutes, you are certainly capable of writing down a formula $f$ that has $300$
variables, call them $p_1, \ldots, p_{300}$. The truth table of $f$ will have $2^{300}$ rows,
which is more than the number of atoms in the observable universe. You can learn a lot more
about {\mc SAT} in a higher-level class on computational complexity (e.g., {\mc COMP 360/362}).

\advsect Predicate logic

\datestamp{5}{}
Propositional logic allows us to work with simple declarations, but this isn't powerful enough
to express some deeper mathematical concepts. For this purpose, we now introduce the notion of
a {\it predicate}. This is a statement involving some number of variables, each of which may take
values coming from a universe $U$, such that the statement evaluates to either true or false {\sl once
all variables are assigned values}. The statement $P(n)$ given by ``$n$ is prime''
is an example of this, where $n$
can take any value in the universe $\ZZ$. An example with two variables is the
predicate $L(x,y)$ defined by ``$x$ is less than $y$.''
(A more commonly-used notation for this predicate is ``$x<y$.'')

Predicates contain variables, but at the moment we don't have any way of introducing new variables into
a statement. This is done using two different {\it quantifiers}.
The first is the
{\it universal quantifier}, denoted $\forall$ and meaning ``for all.'' The statement $\forall n:P(n)$
is true if and only if $P(n)$ is true for every possible value that $n$ can take. The second quantifier
is the {\it existential quantifier}, written ``$\exists$'' and with the meaning ``there exists.''
The statement $\exists n:P(n)$ is true if and only if there is (at least) one value that $n$ can take
such that $P(n)$ is true. The colon doesn't really have any mathematical meaning in these formulas;
they just visually set the quantifiers apart from the predicates that follow.

For instance, taking $P(n)$ to be the statement ``$n$ is prime,'' where the universe $U$ is $\NN$,
the statement $\exists n P(n)$ is true and the statement $\forall n:P(n)$ is false.
What about the statement $\forall x\,\exists y:y< x$? Well, if the universe $U$ is taken to be $\NN$,
then the statement is false, because setting $x$ equal to $0$, there is no element $y$ of $\NN$ such that
$y<0$. But if $U = \ZZ$, then the statement is true, since for every integer $x$, we can put $y = x-1$,
so that $y<x$.

Now we practise translating converting mathematical statements written in English into formulas
in predicate logic. Suppose we want to write: ``Every integer is even or odd.'' The universe here is
is the set $\ZZ$ of integers. The word ``every'' has the same meaning as ``for all,'' so right
off the bat, we can reexpress the statement as: ``For all $n\in \ZZ$, $n$ is even or $n$ is odd.''
In symbols, this is
$$ \forall n\,(n\ \hbox{even} \vee n\ \hbox{odd}).$$
Lastly, we need to figure out how to express the property of being even or being odd.
An integer $n$ is even if an only if it is a multiple of two; that is, if there is some integer $k$
such that $2k = n$. Likewise, an integer is odd if and only if it is one more than a multiple of two.
The corresponding formula is $\exists k : 2k+1 = n$.
So our statement can be expressed
$$\forall n\, \bigl((\exists k : n = 2k)\vee (\exists k : n = 2k+1)\bigr).$$
The variable $k$ appears twice in this formula, but its first instance is independent of its second instance,
because of the parentheses. (Readers who write computer programs will be familiar with the concept of
a variable ``going out of scope.'') So there is nothing wrong with this formula, but to be extra
clear that the first $k$ is different from the second $k$, why don't we replace it with a different letter?
Thus we arrive at
$$\forall n\, \bigl((\exists k : n = 2k)\vee (\exists l : n = 2l+1)\bigr),$$
a formula in predicate logic that means ``every integer is even or odd.''

\medskip\boldlabel Restrictions using quantifiers. It is not true that every real number has a multiplicative
inverse, since one cannot divide by zero. However, the statement ``every nonzero real number has a multiplicative
inverse'' is true. How should we write this as a formula in predicate logic?
One way is to use the conditional: over the universe $U = \RR$, we could write
$$\forall x : (x\ne 0 \To \exists y : xy = 1).$$
Another is to use subscripts: in the same universe, we write
$$\forall x_{x\ne 0}\, \exists y : xy = 1.$$
Using subscripts is slightly informal, since we didn't formally define above what a subscript is supposed
to mean, but it is something that you might encounter. The last way is to simply restrict the universe itself:
in the universe $U = \RR\setminus\{0\}$, the statement
$$\forall x\, \exists y : xy = 1$$
is true.

\medskip\boldlabel Multiple quantifiers. Withing a formula, quantifiers cannot be interchanged
willy-nilly. The order of $\forall$ and $\exists$ matters! They are read from left to right.
Consider the following examples, over the universe $U = \RR$.
The statement
$$ \forall x\, \exists y : x+y = 0$$
is true, since for each given $x$ we can take $y$ to be $-x$. On the other hand,
$$ \exists y\, \forall x : x+y = 0$$
is false, since it would mean that there is some integer that adds up to zero with {\it any} integer.
Of course, sometimes switching the order of quantifiers, doesn't change the truth value of a statement.
Both
$$\exists y\, \forall x: xy = 0$$
and
$$\forall x\, \exists y: xy = 0$$
are true, since in the first case, we can take $y = 0$, and in the second case, we can set $y$ to $0$ no
matter what $x$ is given.

So we know that the order of $\forall$ and $\exists$ matters in general, but repeated instances
of the {\it same} quantifier {\it can} be interchanged. For instance,
$$\forall x\, \forall y : x^2 + y^4 \ge 0$$
is the same as
$$\forall y\, \forall x : x^2 + y^4 \ge 0,$$
and we can even write $\forall x,y : x^2 + y^4 \ge 0$, to introduce both variables simultaneously.

\medskip\boldlabel Negating quantifiers. The universal quantifier is sort of like a big chain of conjunctions
that goes over the whole of the universe. For example, in the universe $\NN$, the
statement $\forall n : P(n)$ is equivalent to
$$P(1) \wedge P(2) \wedge P(3) \wedge \cdots,$$
if this were a valid propositional formula (it isn't because we don't allow propositional formulas to be
infinite). Likewise, the existential quantifier $\exists n : P(n)$ is equivalent to
$$P(1)\vee P(2) \vee P(3) \vee\cdots.$$
We know, by De Morgan's laws, that negating a big series of conjunctions requires us to
flip all the {\mc AND}s to {\mc OR}s. So, once again abusing notation somewhat, we expect
$$\neg\bigl(P(1) \wedge P(2) \wedge P(3) \wedge \cdots\bigr)
\equiv \neg P(1) \vee \neg P(2)\vee \neg P(3)\vee\cdots.$$
Thus we conclude that
$$\neg\bigl(\forall n : P(n)\bigr) \equiv \exists n : \neg P(n).$$
You can play the same game with the other De Morgan's law to show that
$$\neg \bigl(\exists n : P(n)\bigr) \equiv \forall n : \neg P(n).$$
Going back to our example of $P(n)$ denoting ``$n$ is prime,'' the statement $\neg \bigl( \forall n : P(n)\bigr)$
is true, since not all integers $n$ are prime, and we have just shown that this is equivalent to the
statement $\exists n : \neg P(n)$; that is, there exists $n$ such that $n$ is not prime.
